---
title: "Copula Mixed Models"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Copula Mixed Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# pheatmap grobs must be drawn explicitly for knitr figure capture
draw_pheatmap <- function(grob) {
  grid::grid.newpage()
  grid::grid.draw(if (!is.null(grob$gtable)) grob$gtable else grob)
  invisible(NULL)
}
```

## Motivation

Gaussian copula / nonparanormal models map each margin to a latent Gaussian
score \(z_j = \Phi^{-1}(F_j(y_j))\) and then model dependence among the \(z_j\).
In longitudinal, repeated-measures, and clustered studies, observations within
a subject (or cluster) share latent offsets that are **not** part of the
cross-variable copula. Putting random effects in the **observed** margins
conflates marginal location with dependence. This package places them only in
the latent Gaussian layer:

\[
z_j = X\beta_j + Z b_j + \varepsilon_j, \qquad
b_j \sim \mathrm{MVN}(0, D_j), \qquad
\varepsilon \sim \mathrm{MVN}(0, \Sigma).
\]

Here `nodeCols` still defines which variables enter the copula. The shared
RHS-only `latentFormula` (e.g. `~ time + (1 + time | subject)`) specifies the
**latent mean structure** applied to each node's scores. Residual \(\Sigma\)
is the copula covariance; when a meta-analytic NIW prior is available it is
updated with the existing powerprior machinery unchanged.

## Setup

```{r libs}
library(copulaNetworks)
```

This vignette uses `engine = "conjugate"` so it runs without Stan. We recommend
`engine = "brms"` (Hamiltonian Monte Carlo) when **brms** is installed.

Heatmaps and network plots use the same exported `Plot*` helpers as the
factor-vine and meta-analysis vignettes (`pheatmap` and `ggraph`).

## Simulated longitudinal data

```{r simulate}
set.seed(42)
n_subj <- 30
n_time <- 4
d <- 6
subject <- rep(seq_len(n_subj), each = n_time)
time <- rep(seq_len(n_time), times = n_subj)
n <- length(subject)
b0 <- matrix(rnorm(n_subj * d, 0, 0.55), n_subj, d)
b1 <- matrix(rnorm(n_subj * d, 0, 0.12), n_subj, d)
f <- rnorm(n)
lam <- runif(d, 0.35, 0.65)
eps <- matrix(rnorm(n * d), n, d)
X <- matrix(NA_real_, n, d)
for (j in seq_len(d)) {
  X[, j] <- b0[subject, j] + b1[subject, j] * time +
    0.2 * time + lam[j] * f + eps[, j] * sqrt(1 - lam[j]^2)
}
colnames(X) <- paste0("node", seq_len(d))
dat <- cbind(data.frame(subject = subject, time = time), as.data.frame(X))
nodes <- paste0("node", seq_len(d))
```

## Latent mixed fit (random intercept + slope)

```{r fit-mixed}
fit_mixed <- suppressWarnings(FitCopulaLatentMixedModel(
  dat,
  nodeCols = nodes,
  latentFormula = ~ time + (1 + time | subject),
  engine = "conjugate",
  nIter = 400,
  burnIn = 100,
  thin = 2,
  seed = 42
))
fit_mixed$engine
dim(fit_mixed$residualMatrix)
head(fit_mixed$fixedEffects)
```

## Figures: raw NPN vs residual copula (same Plot* suite)

Raw nonparanormal scores (no random-effect adjustment):

```{r raw-cor-heatmap, fig.cap="Copula correlation on raw NPN scores (no RE adjustment)"}
z_raw <- NonparanormalTransform(as.matrix(dat[, nodes]))
raw_fit <- FitStratumCopula(
  as.data.frame(z_raw),
  nodeCols = nodes,
  nlambda = 20,
  method = "ebic",
  preTransformed = TRUE
)
draw_pheatmap(
  PlotCopulaCorHeatmap(raw_fit, title = "Raw NPN copula correlation")
)
```

Residual correlation after latent mixed-model adjustment:

```{r residual-cor-heatmap, fig.cap="Residual copula correlation after latent mixed model"}
draw_pheatmap(
  PlotLatentMixedResidualCorHeatmap(
    fit_mixed,
    title = "Residual copula correlation"
  )
)
```

```{r residual-pcor-heatmap, fig.cap="Residual partial correlation after latent mixed model"}
draw_pheatmap(
  PlotLatentMixedResidualPcorHeatmap(
    fit_mixed,
    title = "Residual partial correlation"
  )
)
```

```{r residual-network, fig.cap="Residual graphical network"}
print(PlotLatentMixedResidualNetwork(
  fit_mixed,
  title = "Residual network",
  printPlot = FALSE
))
```

## Comparison: no-RE vs residual-copula structure

```{r comparison}
resid_graphical <- PseudoGraphicalFitFromLatentMixed(fit_mixed)
cmp <- CompareTwoStrata(
  raw_fit,
  resid_graphical,
  labelA = "rawNPN",
  labelB = "residualMixed",
  deltaThreshold = 0.05
)
cmp_plots <- PlotStratumComparison(cmp)
```

```{r comparison-pcor-scatter, fig.cap="Partial correlation: raw NPN vs residual mixed", fig.width=8, fig.height=6}
print(cmp_plots$plots$pcor$scatter)
```

```{r comparison-cor-scatter, fig.cap="Copula correlation: raw NPN vs residual mixed", fig.width=8, fig.height=6}
print(cmp_plots$plots$copulaCor$scatter)
```

## Repeated measures (random intercept only)

```{r repeated}
set.seed(7)
n_clust <- 40
n_rep <- 3
clust <- rep(seq_len(n_clust), each = n_rep)
n2 <- length(clust)
b <- matrix(rnorm(n_clust * 5, 0, 0.7), n_clust, 5)
f2 <- rnorm(n2)
lam2 <- runif(5, 0.3, 0.55)
Y <- matrix(NA_real_, n2, 5)
for (j in 1:5) {
  Y[, j] <- b[clust, j] + lam2[j] * f2 + rnorm(n2) * sqrt(1 - lam2[j]^2)
}
colnames(Y) <- paste0("m", 1:5)
rep_dat <- cbind(data.frame(cluster = clust), as.data.frame(Y))
rep_nodes <- paste0("m", 1:5)

fit_rep <- suppressWarnings(FitCopulaLatentMixedModel(
  rep_dat,
  nodeCols = rep_nodes,
  latentFormula = ~ 1 + (1 | cluster),
  engine = "conjugate",
  nIter = 300,
  burnIn = 80,
  seed = 7
))
```

```{r repeated-cor-heatmap, fig.cap="Repeated-measures residual copula correlation"}
draw_pheatmap(
  PlotLatentMixedResidualCorHeatmap(
    fit_rep,
    title = "Repeated measures: residual correlation"
  )
)
```

```{r repeated-network, fig.cap="Repeated-measures residual network"}
print(PlotLatentMixedResidualNetwork(
  fit_rep,
  title = "Repeated measures: residual network",
  printPlot = FALSE
))
```

## Interpreting fixed effects, random effects, and Sigma

- **Fixed effects** (`fit$fixedEffects`): population-level associations on the
  latent scale (e.g. time trend shared across subjects).
- **Random effects** (`fit$randomEffects`, `fit$randomCov`): subject-specific
  latent intercepts/slopes with node-specific \(D_j\).
- **Residual \(\Sigma\)**: remaining cross-node dependence after removing
  within-subject latent structure -- this is the copula network of interest.

Prediction on the latent scale:

```{r predict}
pred <- PredictCopulaLatentMixedModel(fit_mixed, dat, type = "latentMean")
dim(pred)
```

## Optional residual NIW update

When historical cohorts define a [FitMetaCorPrior()] prior, residual scores
feed the existing NIW posterior (unchanged conjugacy):

```{r niw}
hist <- list(
  A = as.data.frame(matrix(rnorm(120 * d), 120, d, dimnames = list(NULL, nodes))),
  B = as.data.frame(matrix(rnorm(100 * d), 100, d, dimnames = list(NULL, nodes)))
)
meta_prior <- FitMetaCorPrior(hist, nodeCols = nodes, a0 = 0.5)
fit_niw <- suppressWarnings(FitCopulaLatentMixedModel(
  dat,
  nodeCols = nodes,
  latentFormula = ~ time + (1 | subject),
  engine = "conjugate",
  metaPrior = meta_prior,
  nIter = 200,
  burnIn = 50,
  seed = 1
))
draw_pheatmap(
  PlotCopulaCorHeatmap(
    list(copulaCor = fit_niw$impliedCor, keptCols = nodes),
    title = "Posterior residual correlation (NIW)"
  )
)
```

## Relation to existing workflows

| Workflow | Entry point | Role of random effects |
|----------|-------------|------------------------|
| Stratified glasso | `RunCopulaPipeline()` | None (NPN scores directly) |
| Factor-vine | `RunFactorVinePipeline()` | None (factor prior + update) |
| Meta power prior | `RunMetaAnalysisPipeline()` | None (NIW on scores) |
| **Latent mixed** | `FitCopulaLatentMixedModel()` | RE in latent layer; NIW/glasso on residuals |

Use the latent mixed model when clustering / longitudinal structure would
otherwise inflate or distort the copula network. Users who omit
`latentFormula` keep the existing no-RE workflows unchanged.

## Session info

```{r session}
sessionInfo()
```
