---
title: "Bayesian Meta-Analysis with Power Prior"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Meta-Analysis with Power Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

draw_pheatmap <- function(grob) {
  grid::grid.newpage()
  grid::grid.draw(if (!is.null(grob$gtable)) grob$gtable else grob)
  invisible(NULL)
}
```

## Motivation

When several **large historical cohorts** inform dependence among the same
variables, a Bayesian **meta-analysis** can pool them before updating with a
**small** new sample. The **power prior** (Ibrahim & Chen, 2000) discounts each
cohort's contribution via \(a_0 \in [0,1]\), controlling borrowing strength.
Zhang et al. (2017) motivate study-specific discounting for correlation
meta-analysis; here we use the multivariate Normal-Inverse-Wishart conjugate
framework in the [**powerprior**](https://cran.r-project.org/package=powerprior)
package on Gaussian copula normal scores.

## Setup

```{r libs}
library(copulaNetworks)
```

```{r check-deps}
has_powerprior <- requireNamespace("powerprior", quietly = TRUE)
```

| Package | Role |
|---------|------|
| powerprior | Multivariate NIW power prior and posterior |
| huge | Phase 2 graphical lasso (imported) |

## Synthetic historical cohorts

Three large cohorts share a latent factor structure with mild heterogeneity:

```{r synthetic}
set.seed(42)
d <- 10
n_update <- 30
lambda <- runif(d, 0.35, 0.7)

sim_block <- function(n_obs, lambda_jitter = 0) {
  lam <- pmax(0.2, pmin(0.85, lambda + lambda_jitter))
  f <- rnorm(n_obs)
  eps <- matrix(rnorm(n_obs * d), nrow = n_obs, ncol = d)
  factor_part <- matrix(lam, nrow = n_obs, ncol = d, byrow = TRUE) * f
  x <- factor_part + sweep(eps, 2, sqrt(1 - lam^2), `*`)
  colnames(x) <- paste0("node", seq_len(d))
  as.data.frame(x)
}

historicalCohorts <- list(
  cohortA = sim_block(400, lambda_jitter = 0),
  cohortB = sim_block(350, lambda_jitter = 0.03),
  cohortC = sim_block(300, lambda_jitter = -0.02)
)
updateData <- sim_block(n_update)
nodes <- paste0("node", 1:d)
lapply(historicalCohorts, nrow)
```

## Per-cohort correlation heatmaps

```{r cohort-heatmaps, eval=has_powerprior, fig.cap="Per-cohort copula correlation (empirical on normal scores)"}
for (cohort_name in names(historicalCohorts)) {
  cohort_df <- historicalCohorts[[cohort_name]]
  z_cohort <- qnorm(UniformMarginalTransform(as.matrix(cohort_df)))
  cohort_cor <- cor(z_cohort)
  draw_pheatmap(
    PlotCopulaCorHeatmap(
      list(copulaCor = cohort_cor, keptCols = nodes),
      title = paste("Cohort:", cohort_name)
    )
  )
}
```

## Phase 1: Meta-analytic power prior

```{r phase1, eval=has_powerprior}
meta_prior <- FitMetaCorPrior(
  historicalCohorts,
  nodeCols = nodes,
  a0 = c(0.6, 0.5, 0.5)
)
meta_prior$cohortSummaries
meta_prior$nHistorical
```

Meta-analytic implied correlation:

```{r phase1-heatmap, eval=has_powerprior, fig.cap="Phase 1 meta-analytic implied correlation"}
draw_pheatmap(
  PlotCopulaCorHeatmap(
    list(copulaCor = meta_prior$impliedCor, keptCols = nodes),
    title = "Meta-analytic prior correlation"
  )
)
```

Prior network (dense pseudo-graphical fit from implied correlation):

```{r phase1-network, eval=has_powerprior, fig.cap="Meta-analytic prior network"}
prior_pseudo <- PseudoGraphicalFitFromCor(meta_prior$impliedCor, nObs = meta_prior$n)
print(PlotCopulaNetwork(
  prior_pseudo,
  title = "Meta-analytic prior",
  printPlot = FALSE
))
```

## Phase 2: Graphical update on small sample

Update data use **pooled historical** marginal transforms:

```{r phase2, eval=has_powerprior}
update_fit <- FitCopulaUpdate(
  updateData,
  priorFit = meta_prior,
  method = "graphical",
  nlambda = 20
)
update_fit$graphical$lambdaOpt
```

```{r phase2-cor-heatmap, eval=has_powerprior, fig.cap="Phase 2 update copula correlation"}
draw_pheatmap(
  PlotCopulaCorHeatmap(
    update_fit$graphical,
    title = "Phase 2 update copula correlation"
  )
)
```

```{r phase2-pcor-heatmap, eval=has_powerprior, fig.cap="Phase 2 update partial correlation"}
draw_pheatmap(
  PlotPcorHeatmap(
    update_fit$graphical,
    title = "Phase 2 update partial correlation"
  )
)
```

```{r phase2-network, eval=has_powerprior, fig.cap="Phase 2 update graphical network"}
print(PlotCopulaNetwork(
  update_fit$graphical,
  title = "Phase 2 update",
  printPlot = FALSE
))
```

## Prior vs update comparison

```{r comparison, eval=has_powerprior}
cmp <- ComparePriorToUpdate(meta_prior, update_fit)
cmp_plots <- PlotStratumComparison(cmp)
```

```{r comparison-pcor-scatter, eval=has_powerprior, fig.cap="Partial correlation: meta prior vs update", fig.width=8, fig.height=6}
print(cmp_plots$plots$pcor$scatter)
```

```{r comparison-cor-scatter, eval=has_powerprior, fig.cap="Copula correlation: meta prior vs update", fig.width=8, fig.height=6}
print(cmp_plots$plots$copulaCor$scatter)
```

## Bayesian conjugate update

```{r bayes, eval=has_powerprior}
bayes_fit <- FitBayesianMetaUpdate(meta_prior, updateData)
head(bayes_fit$impliedCor[, 1:4])
```

Posterior correlation heatmap:

```{r bayes-heatmap, eval=has_powerprior, fig.cap="Posterior correlation after powerprior update"}
draw_pheatmap(
  PlotCopulaCorHeatmap(
    list(copulaCor = bayes_fit$impliedCor, keptCols = nodes),
    title = "Posterior correlation (conjugate update)"
  )
)
```

## Discount sensitivity

Global discount grid (same `a0` recycled across cohorts):

```{r sensitivity, eval=has_powerprior}
sens <- RunDiscountSensitivity(
  meta_prior,
  updateData,
  a0Values = c(0, 0.25, 0.5, 0.75, 1)
)
sens$summary
```

Univariate diagnostic via **powerprior** (first node mean on pooled historical Z):

```{r sensitivity-plot, eval=has_powerprior, fig.cap="Discount sensitivity (powerprior)"}
z_pooled <- do.call(rbind, meta_prior$zCohortList)
z_update <- bayes_fit$zMatrix
cmp_disc <- powerprior::compare_discounting(
  z_pooled,
  z_update,
  a0_values = c(0, 0.25, 0.5, 0.75, 1),
  multivariate = TRUE
)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(powerprior::plot_sensitivity(cmp_disc, parameter = "mu", dimension = 1))
}
```

Posterior correlation at selected discount values:

```{r sensitivity-cor-heatmaps, eval=has_powerprior, fig.cap="Posterior correlation vs global a0"}
for (a0_key in c("0", "0.5", "1")) {
  draw_pheatmap(
    PlotCopulaCorHeatmap(
      list(copulaCor = sens$corMatrices[[a0_key]], keptCols = nodes),
      title = paste("Posterior cor, global a0 =", a0_key)
    )
  )
}
```

Compatibility of each historical cohort with the update sample:

```{r compatibility, eval=has_powerprior}
compat <- lapply(names(meta_prior$zCohortList), function(nm) {
  powerprior::check_compatibility(meta_prior$zCohortList[[nm]], bayes_fit$zMatrix)
})
names(compat) <- names(meta_prior$zCohortList)
compat
```

## End-to-end pipeline

```{r pipeline, eval=has_powerprior}
res <- RunMetaAnalysisPipeline(
  historicalCohorts = historicalCohorts,
  updateData = updateData,
  nodeCols = nodes,
  a0 = c(0.6, 0.5, 0.5),
  a0Sensitivity = c(0, 0.5, 1),
  phase2Method = "graphical",
  nlambda = 20
)
res$bayesFit$impliedCor[1, 2]
nrow(res$sensitivity$summary)
```

## Relation to factor-vine workflow

Use [`RunFactorVinePipeline()`](https://GWMcElfresh.github.io/copulaNetworks/reference/RunFactorVinePipeline.html)
when a **single** large reference cohort supports a factor copula prior.
Use `RunMetaAnalysisPipeline()` when **multiple** historical cohorts should be
pooled with study-specific discounting before a small-sample update.

## Limitations

- Sequential NIW chaining (not a full multi-study product likelihood)
- Global `a0` only in `RunDiscountSensitivity()`; per-cohort `a0` in `FitMetaCorPrior()`
- Correlation from NIW mean covariance, not full posterior over correlation elements
- Random-effects meta-analysis (Zhang et al. §2.2) deferred
- Covariate-adjusted marginals: supply pre-adjusted data in v1

## References

- Ibrahim, J. G., & Chen, M. H. (2000). Power prior distributions for regression models. *Statistical Science*, 15(1), 46–60.
- Zhang, Z., Jiang, K., Liu, H., & Oh, I.-S. (2017). Bayesian meta-analysis of correlation coefficients through power prior. *Communications in Statistics - Theory and Methods*, 46(24), 11988–12007. [doi:10.1080/03610926.2017.1288251](https://doi.org/10.1080/03610926.2017.1288251)

## Session info

```{r session}
sessionInfo()
```
