---
title: "Stratified Nonparanormal Copula Networks"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stratified Nonparanormal Copula Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# pheatmap grobs must be drawn explicitly for knitr figure capture
draw_pheatmap <- function(grob) {
  grid::grid.newpage()
  grid::grid.draw(if (!is.null(grob$gtable)) grob$gtable else grob)
  invisible(NULL)
}
```

## Motivation

Understanding dependence among continuous biomarkers or imaging features is
central to network-style analyses, but the variables rarely share a common
scale or a Gaussian margin. Pearson correlation on the raw measurements
therefore confounds marginal shape with association. The nonparanormal
(Gaussian copula) model maps each margin to a latent normal score
\(z_j = \Phi^{-1}(\hat{F}_j(y_j))\) and then estimates dependence among the
\(z_j\). Graphical lasso on those scores targets a sparse precision graph in
the copula space rather than in the original units.

This vignette is the base workflow for **copulaNetworks**: transform, fit a
glasso network (StARS or eBIC), and visualize correlations and edges. Later
articles extend the same normal-score layer with factor priors, meta-analytic
power priors, and latent random effects.

## Setup

```{r libs}
library(copulaNetworks)
```

Heatmaps and network plots use the exported `Plot*` helpers (`pheatmap` and
`ggraph`). We keep `nlambda` modest so the vignette builds quickly; production
fits often use a denser path (e.g. `nlambda = 40`).

## Single-stratum warm-up

A one-factor residual model produces dense correlations but nearly empty
partial-correlation graphs after glasso, because most pairs are explained by
the shared factor. To induce recoverable edges we simulate from a **sparse
Gaussian graphical model**: a chain plus a few long-range edges in the
precision matrix \(\Omega\), then apply mild nonlinear margin transforms so
the observed scale is non-Gaussian while dependence lives in the latent
scores.

```{r synthetic-single}
set.seed(42)
d <- 8
n_obs <- 200
nodes <- paste0("node", seq_len(d))

# Sparse precision: chain 1-2-...-d plus hubs 1-5 and 3-7. Off-diagonals set
# the target partial-correlation scale; a small eigen ridge enforces PD.
make_sparse_precision <- function(d,
                                  edge_strength = 0.6,
                                  extra_edges = list(c(1L, 5L), c(3L, 7L))) {
  omega <- diag(d)
  for (i in seq_len(d - 1L)) {
    omega[i, i + 1L] <- omega[i + 1L, i] <- -edge_strength
  }
  for (edge in extra_edges) {
    i <- edge[1]
    j <- edge[2]
    omega[i, j] <- omega[j, i] <- -0.5
  }
  ev_min <- min(eigen(omega, symmetric = TRUE, only.values = TRUE)$values)
  if (ev_min <= 0) {
    omega <- omega + diag(abs(ev_min) + 0.2, d)
  }
  omega
}

sim_nonparanormal_ggm <- function(n_obs, omega, warp = TRUE) {
  sigma <- solve(omega)
  z <- matrix(rnorm(n_obs * nrow(omega)), n_obs, nrow(omega)) %*% chol(sigma)
  colnames(z) <- nodes
  if (!isTRUE(warp)) {
    return(as.data.frame(z))
  }
  # Rank-preserving warps: nonparanormal should still recover the graph.
  x <- z
  x[, 1] <- exp(z[, 1] / 2)
  x[, 2] <- z[, 2]^3
  x[, 4] <- sinh(z[, 4])
  as.data.frame(x)
}

omega_full <- make_sparse_precision(d, edge_strength = 0.6)
dat <- sim_nonparanormal_ggm(n_obs, omega_full)
dim(dat)
true_edges <- sum(abs(omega_full[upper.tri(omega_full)]) > 1e-8)
true_edges
```

Fit one nonparanormal graphical model. StARS selects \(\lambda\) by edge
stability; with the default threshold `0.1` this simulation is often too
conservative to show the chain, so the vignette uses `starsThresh = 0.2` to
keep a few recoverable edges visible.

```{r fit-single}
fit_full <- FitStratumCopula(
  dat,
  nodeCols = nodes,
  method = "stars",
  nlambda = 40,
  starsThresh = 0.2
)
fit_full$n
fit_full$lambdaOpt
sum(fit_full$adjacency[upper.tri(fit_full$adjacency)])
```

Copula correlation (empirical on normal scores):

```{r single-cor, fig.cap="Single-stratum copula correlation"}
draw_pheatmap(
  PlotCopulaCorHeatmap(fit_full, title = "Copula correlation (full sample)")
)
```

Partial correlation after glasso:

```{r single-pcor, fig.cap="Single-stratum partial correlation"}
draw_pheatmap(
  PlotPcorHeatmap(fit_full, title = "Partial correlation (glasso)")
)
```

Selected network:

```{r single-network, fig.cap="Single-stratum copula network"}
print(PlotCopulaNetwork(fit_full, title = "Nonparanormal network (full sample)"))
```

## Stratified path

Dependence often differs by cohort, tissue, or timepoint. Both strata share the
chain skeleton; stratum B uses stronger edge weights and adds the edge
\(2\text{--}6\), so comparison can separate shared from stratum-specific
partial correlations. The prepare step only builds named subsets; it does not
impute or transform margins.

```{r synthetic-strata}
set.seed(42)
n_per <- 120
omega_a <- make_sparse_precision(
  d,
  edge_strength = 0.45,
  extra_edges = list(c(1L, 5L), c(3L, 7L))
)
omega_b <- make_sparse_precision(
  d,
  edge_strength = 0.65,
  extra_edges = list(c(1L, 5L), c(3L, 7L), c(2L, 6L))
)
dat_strata <- rbind(
  cbind(group = "A", sim_nonparanormal_ggm(n_per, omega_a)),
  cbind(group = "B", sim_nonparanormal_ggm(n_per, omega_b))
)
dat_strata$group <- as.character(dat_strata$group)
table(dat_strata$group)
```

```{r prepare}
prep <- PrepareCopulaData(
  data = dat_strata,
  strataCols = "group",
  nodeCols = nodes,
  strataSpecs = list(
    group = list(
      group_by = "group",
      minN = 30
    )
  )
)
prep$meta
names(prep$strata)
```

```{r fit-strata}
fits <- FitCopulaStrata(
  prep,
  method = "stars",
  nlambda = 40,
  starsThresh = 0.2,
  minN = 30,
  includeFull = FALSE
)
names(fits$fits)
vapply(
  fits$fits,
  function(fit) sum(fit$adjacency[upper.tri(fit$adjacency)]),
  numeric(1)
)
```

Per-stratum correlation heatmaps:

```{r strata-heatmaps, fig.cap="Per-stratum copula correlation"}
for (nm in names(fits$fits)) {
  draw_pheatmap(
    PlotCopulaCorHeatmap(
      fits$fits[[nm]],
      title = paste("Copula correlation:", nm)
    )
  )
}
```

Per-stratum networks:

```{r strata-networks, fig.cap="Per-stratum copula networks"}
for (nm in names(fits$fits)) {
  print(PlotCopulaNetwork(
    fits$fits[[nm]],
    title = paste("Network:", nm)
  ))
}
```

When both strata fit, pairwise comparison separates shared edges from
stratum-specific structure:

```{r compare}
nm_a <- "group::A"
nm_b <- "group::B"
cmp <- CompareTwoStrata(
  fits$fits[[nm_a]],
  fits$fits[[nm_b]],
  labelA = "A",
  labelB = "B",
  matrices = c("pcor", "copulaCor")
)
cmp_plots <- PlotStratumComparison(cmp)
```

```{r compare-pcor, fig.cap="Partial correlation: stratum A vs B", fig.width=8, fig.height=6}
print(cmp_plots$plots$pcor$scatter)
```

## One-call pipeline

`RunCopulaPipeline()` chains prepare, fit, optional diagnostics, and
comparisons, writing RDS checkpoints when `outDir` is set. Diagnostics and
pairwise plot exports are omitted here because those steps were already run
above; production calls typically set `plotDiagnostics = TRUE` and pass
`comparePairs`.

```{r pipeline}
out_dir <- file.path(tempdir(), "copula_strata_vignette")
pipeline_res <- RunCopulaPipeline(
  data = dat_strata,
  strataCols = "group",
  nodeCols = nodes,
  strataSpecs = list(
    group = list(group_by = "group", minN = 30)
  ),
  outDir = out_dir,
  method = "stars",
  nlambda = 40,
  starsThresh = 0.2,
  minN = 30,
  plotDiagnostics = FALSE
)
names(pipeline_res$fitResults$fits)
```

## Where next

The stratified nonparanormal fit is the package baseline: normal scores, glasso,
and stratum comparison. Three extensions build on the same latent Gaussian
layer:

- **Two-phase factor-vine** — large-sample factor prior, then small-sample
  graphical or vine update.
- **Meta-analysis power prior** — pool several historical cohorts with cohort-
  specific discounting before a small-sample update.
- **Latent mixed models** — place random effects in the latent scores so
  longitudinal or clustered structure does not inflate residual dependence.

## Session info

```{r session}
sessionInfo()
```
