---
title: "Stratified Nonparanormal Copula Networks"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stratified Nonparanormal Copula Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# pheatmap grobs must be drawn explicitly for knitr figure capture
draw_pheatmap <- function(grob) {
  grid::grid.newpage()
  grid::grid.draw(if (!is.null(grob$gtable)) grob$gtable else grob)
  invisible(NULL)
}
```

## Motivation

Understanding dependence among continuous biomarkers or imaging features is
central to network-style analyses, but the variables rarely share a common
scale or a Gaussian margin. Pearson correlation on the raw measurements
therefore confounds marginal shape with association. The nonparanormal
(Gaussian copula) model maps each margin to a latent normal score
\(z_j = \Phi^{-1}(\hat{F}_j(y_j))\) and then estimates dependence among the
\(z_j\). Graphical lasso on those scores targets a sparse precision graph in
the copula space rather than in the original units.

This vignette is the base workflow for **copulaNetworks**: transform, fit a
glasso network (StARS or eBIC), and visualize correlations and edges. Later
articles extend the same normal-score layer with factor priors, meta-analytic
power priors, and latent random effects.

## Setup

```{r libs}
library(copulaNetworks)
```

Heatmaps and network plots use the exported `Plot*` helpers (`pheatmap` and
`ggraph`). We keep `nlambda` modest so the vignette builds quickly; production
fits often use a denser path (e.g. `nlambda = 40`).

## Single-stratum warm-up

Simulate a continuous block with one shared latent factor and idiosyncratic
noise. Loadings in \([0.35, 0.75]\) keep residual variance identifiable while
inducing a dense correlation skeleton for glasso to sparsify.

```{r synthetic-single}
set.seed(42)
d <- 8
n_obs <- 120
lambda <- runif(d, 0.35, 0.75)
f <- rnorm(n_obs)
eps <- matrix(rnorm(n_obs * d), nrow = n_obs, ncol = d)
factor_part <- matrix(lambda, nrow = n_obs, ncol = d, byrow = TRUE) * f
x <- factor_part + sweep(eps, 2, sqrt(1 - lambda^2), `*`)
colnames(x) <- paste0("node", seq_len(d))
dat <- as.data.frame(x)
nodes <- colnames(dat)
dim(dat)
```

Fit one nonparanormal graphical model. StARS selects \(\lambda\) by edge
stability rather than by information criteria alone, which is useful when
\(n\) is moderate relative to \(p\).

```{r fit-single}
fit_full <- FitStratumCopula(
  dat,
  nodeCols = nodes,
  method = "stars",
  nlambda = 20,
  starsThresh = 0.1
)
fit_full$n
fit_full$lambdaOpt
```

Copula correlation (empirical on normal scores):

```{r single-cor, fig.cap="Single-stratum copula correlation"}
draw_pheatmap(
  PlotCopulaCorHeatmap(fit_full, title = "Copula correlation (full sample)")
)
```

Partial correlation after glasso:

```{r single-pcor, fig.cap="Single-stratum partial correlation"}
draw_pheatmap(
  PlotPcorHeatmap(fit_full, title = "Partial correlation (glasso)")
)
```

Selected network:

```{r single-network, fig.cap="Single-stratum copula network"}
print(PlotCopulaNetwork(fit_full, title = "Nonparanormal network (full sample)"))
```

## Stratified path

Dependence often differs by cohort, tissue, or timepoint. Here we add a binary
group with a mild loading shift so stratum B is slightly more factor-driven than
stratum A. The prepare step only builds named subsets; it does not impute or
transform margins.

```{r synthetic-strata}
set.seed(42)
n_per <- 80
sim_group <- function(n_obs, lambda_shift = 0) {
  lam <- pmax(0.2, pmin(0.85, lambda + lambda_shift))
  f <- rnorm(n_obs)
  eps <- matrix(rnorm(n_obs * d), nrow = n_obs, ncol = d)
  factor_part <- matrix(lam, nrow = n_obs, ncol = d, byrow = TRUE) * f
  x <- factor_part + sweep(eps, 2, sqrt(1 - lam^2), `*`)
  colnames(x) <- nodes
  as.data.frame(x)
}
dat_strata <- rbind(
  cbind(group = "A", sim_group(n_per, lambda_shift = -0.05)),
  cbind(group = "B", sim_group(n_per, lambda_shift = 0.08))
)
dat_strata$group <- as.character(dat_strata$group)
table(dat_strata$group)
```

```{r prepare}
prep <- PrepareCopulaData(
  data = dat_strata,
  strataCols = "group",
  nodeCols = nodes,
  strataSpecs = list(
    group = list(
      group_by = "group",
      minN = 30
    )
  )
)
prep$meta
names(prep$strata)
```

```{r fit-strata}
fits <- FitCopulaStrata(
  prep,
  method = "stars",
  nlambda = 20,
  starsThresh = 0.1,
  minN = 30,
  includeFull = FALSE
)
names(fits$fits)
```

Per-stratum correlation heatmaps:

```{r strata-heatmaps, fig.cap="Per-stratum copula correlation"}
for (nm in names(fits$fits)) {
  draw_pheatmap(
    PlotCopulaCorHeatmap(
      fits$fits[[nm]],
      title = paste("Copula correlation:", nm)
    )
  )
}
```

Per-stratum networks:

```{r strata-networks, fig.cap="Per-stratum copula networks"}
for (nm in names(fits$fits)) {
  print(PlotCopulaNetwork(
    fits$fits[[nm]],
    title = paste("Network:", nm)
  ))
}
```

When both strata fit, pairwise comparison separates shared edges from
stratum-specific structure:

```{r compare}
nm_a <- "group::A"
nm_b <- "group::B"
cmp <- CompareTwoStrata(
  fits$fits[[nm_a]],
  fits$fits[[nm_b]],
  labelA = "A",
  labelB = "B",
  matrices = c("pcor", "copulaCor")
)
cmp_plots <- PlotStratumComparison(cmp)
```

```{r compare-pcor, fig.cap="Partial correlation: stratum A vs B", fig.width=8, fig.height=6}
print(cmp_plots$plots$pcor$scatter)
```

## One-call pipeline

`RunCopulaPipeline()` chains prepare, fit, optional diagnostics, and
comparisons, writing RDS checkpoints when `outDir` is set. Diagnostics and
pairwise plot exports are omitted here because those steps were already run
above; production calls typically set `plotDiagnostics = TRUE` and pass
`comparePairs`.

```{r pipeline}
out_dir <- file.path(tempdir(), "copula_strata_vignette")
pipeline_res <- RunCopulaPipeline(
  data = dat_strata,
  strataCols = "group",
  nodeCols = nodes,
  strataSpecs = list(
    group = list(group_by = "group", minN = 30)
  ),
  outDir = out_dir,
  method = "stars",
  nlambda = 20,
  minN = 30,
  plotDiagnostics = FALSE
)
names(pipeline_res$fitResults$fits)
```

## Where next

The stratified nonparanormal fit is the package baseline: normal scores, glasso,
and stratum comparison. Three extensions build on the same latent Gaussian
layer:

- **Two-phase factor-vine** — large-sample factor prior, then small-sample
  graphical or vine update.
- **Meta-analysis power prior** — pool several historical cohorts with cohort-
  specific discounting before a small-sample update.
- **Latent mixed models** — place random effects in the latent scores so
  longitudinal or clustered structure does not inflate residual dependence.

## Session info

```{r session}
sessionInfo()
```
