---
title: "Two-Phase Factor-Vine Copula Networks"
author: "GW McElfresh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Phase Factor-Vine Copula Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# pheatmap grobs must be drawn explicitly for knitr figure capture
draw_pheatmap <- function(grob) {
  grid::grid.newpage()
  grid::grid.draw(if (!is.null(grob$gtable)) grob$gtable else grob)
  invisible(NULL)
}
```

## Motivation

Modern copula models use latent factors to handle high-dimensional dependence.
In a **two-phase** workflow we fit a parsimonious factor copula on a **large**
reference cohort (Phase 1), then update with a graphical model or vine copula on
a **small** new sample (Phase 2), testing whether the update is consistent with
the prior. Factor copulas reduce parameter count in high dimension; vine or
graphical updates refine dependence on small samples; consistency testing checks
whether the update departs from the prior.

## Setup

```{r libs}
library(copulaNetworks)
```

Heatmaps and network plots use the exported `Plot*` helpers (`pheatmap` and
`ggraph`).

Optional dependencies (install as needed):

| Package | Role |
|---------|------|
| FactorCopula | Phase 1 factor copula MLE |
| VineCopula | Phase 2 vine update |
| cmdstanr | Bayesian Gaussian update |
| bridgesampling | Marginal likelihood / Bayes factor |

```{r check-deps}
has_factor <- requireNamespace("FactorCopula", quietly = TRUE)
has_vine <- requireNamespace("VineCopula", quietly = TRUE)
has_cmdstanr <- requireNamespace("cmdstanr", quietly = TRUE)
```

## Synthetic data

```{r synthetic}
set.seed(42)
d <- 12
N <- 500
n <- 30
lambda <- runif(d, 0.35, 0.75)
sim_block <- function(n_obs) {
  f <- rnorm(n_obs)
  eps <- matrix(rnorm(n_obs * d), nrow = n_obs, ncol = d)
  factor_part <- matrix(lambda, nrow = n_obs, ncol = d, byrow = TRUE) * f
  x <- factor_part + sweep(eps, 2, sqrt(1 - lambda^2), `*`)
  colnames(x) <- paste0("node", seq_len(d))
  as.data.frame(x)
}
priorData <- sim_block(N)
updateData <- sim_block(n)
nodes <- paste0("node", 1:d)
dim(priorData)
dim(updateData)
```

## Marginal transforms

Phase 2 must use **prior-fitted** marginal transforms, not re-ranking on small `n`.

```{r marginals}
marginalSpec <- FitMarginalSpec(priorData, nodeCols = nodes)
U_update <- ApplyMarginalSpec(updateData, marginalSpec)
range(U_update)
```

Contrast with re-ranking on the small sample (wrong for copula update):

```{r wrong-rank}
U_wrong <- UniformMarginalTransform(as.matrix(updateData))
mean(abs(U_update - U_wrong))
```

Normal scores for graphical update:

```{r normal-scores}
Z_update <- qnorm(U_update)
head(Z_update[, 1:3])
```

## Phase 1: Factor structure and prior fit

```{r phase1-check, eval=has_factor}
prior_fit <- FitFactorCopulaPrior(
  priorData,
  nodeCols = nodes,
  nFactors = 1L,
  linkingCopula = "bvn"
)
CheckFactorStructure(prior_fit$impliedCor, nFactors = 1L)
prior_fit$loadings
```

Factor-loading groups color nodes in the network plots below:

```{r phase1-node-groups, eval=has_factor}
factor_groups <- FactorGroupsFromLoadings(prior_fit$loadings, k = 2)
node_groups <- setNames(paste0("Group", factor_groups), names(factor_groups))
factor_groups
```

The Phase 1 factor model implies a **dense** correlation structure (every pair
linked through the latent factor). Implied correlation heatmap:

```{r phase1-heatmap, eval=has_factor, fig.cap="Phase 1 factor-implied correlation"}
draw_pheatmap(
  PlotCopulaCorHeatmap(
    list(copulaCor = prior_fit$impliedCor, keptCols = nodes),
    title = "Phase 1 implied correlation"
  )
)
```

Prior implied network (pseudo-graphical fit from implied correlation):

```{r phase1-network, eval=has_factor, fig.cap="Phase 1 factor-implied network"}
prior_pseudo <- PseudoGraphicalFitFromCor(prior_fit$impliedCor, nObs = prior_fit$n)
print(PlotCopulaNetwork(
  prior_pseudo,
  title = "Phase 1 prior (implied)",
  nodeGroups = node_groups,
  printPlot = FALSE
))
```

## Phase 2(a): Vine update

```{r phase2-vine, eval=has_factor && has_vine}
update_fit_both <- FitCopulaUpdate(
  updateData,
  priorFit = prior_fit,
  method = "both",
  nlambda = 20
)
if (!is.null(update_fit_both$vine)) {
  update_fit_both$vine$logLik
}
```

## Phase 2(b): Graphical update

```{r phase2-graphical, eval=has_factor}
update_fit <- FitCopulaUpdate(
  updateData,
  priorFit = prior_fit,
  method = "graphical",
  nlambda = 20
)
update_fit$graphical$lambdaOpt
```

On the small update sample, glasso yields a **sparse** graphical model. Copula
correlation heatmap:

```{r phase2-cor-heatmap, eval=has_factor, fig.cap="Phase 2 update copula correlation"}
draw_pheatmap(
  PlotCopulaCorHeatmap(
    update_fit$graphical,
    title = "Phase 2 update copula correlation"
  )
)
```

Partial correlation heatmap (glasso estimate):

```{r phase2-pcor-heatmap, eval=has_factor, fig.cap="Phase 2 update partial correlation"}
draw_pheatmap(
  PlotPcorHeatmap(
    update_fit$graphical,
    title = "Phase 2 update partial correlation"
  )
)
```

Sparse update network:

```{r network-plot, eval=has_factor, fig.cap="Phase 2 update graphical network"}
print(PlotCopulaNetwork(
  update_fit$graphical,
  title = "Phase 2 update",
  nodeGroups = node_groups,
  printPlot = FALSE
))
```

## Prior vs update comparison

```{r comparison, eval=has_factor}
cmp <- ComparePriorToUpdate(prior_fit, update_fit)
head(cmp$pcor)
cmp_plots <- PlotStratumComparison(cmp)
```

Partial-correlation scatter (prior implied vs update glasso):

```{r comparison-pcor-scatter, eval=has_factor, fig.cap="Partial correlation: prior vs update", fig.width=8, fig.height=6}
print(cmp_plots$plots$pcor$scatter)
```

Largest differential partial-correlation pairs:

```{r comparison-pcor-bar, eval=has_factor, fig.cap="Differential partial correlation pairs", fig.width=8, fig.height=6}
if (!is.null(cmp_plots$plots$pcor$bar)) {
  print(cmp_plots$plots$pcor$bar)
}
```

Copula-correlation scatter (prior implied vs update empirical):

```{r comparison-cor-scatter, eval=has_factor, fig.cap="Copula correlation: prior vs update", fig.width=8, fig.height=6}
print(cmp_plots$plots$copulaCor$scatter)
```

## Frequentist consistency test

```{r gof, eval=has_factor}
gof <- TestPriorConsistency(
  prior_fit,
  updateData,
  nRep = 200,
  seed = 1
)
gof$pValue
hist(gof$nullDistribution, main = "Null distribution", xlab = "pairwise tau stat")
abline(v = gof$obsStat, col = "red", lwd = 2)
```

## Bayesian update

```{r bayes, eval=has_factor && has_cmdstanr, message=FALSE, warning=FALSE}
bayes_fit <- tryCatch(
  FitBayesianFactorUpdate(
    prior_fit,
    updateData,
    chains = 2,
    iter = 500,
    computeBayesFactor = requireNamespace("bridgesampling", quietly = TRUE),
    seed = 1
  ),
  error = function(e) {
    message("Bayesian update skipped: ", conditionMessage(e))
    NULL
  }
)
if (!is.null(bayes_fit)) {
  head(bayes_fit$summary)
}
```

## End-to-end pipeline

```{r pipeline, eval=has_factor}
res <- RunFactorVinePipeline(
  priorData = priorData,
  updateData = updateData,
  nodeCols = nodes,
  nFactors = 1L,
  phase2Method = "graphical",
  testMethod = "simulation",
  nRep = 100,
  nlambda = 20,
  seed = 1
)
res$consistencyTest$pValue
```

Checkpoints use `SaveCheckpoint()` / `LoadCheckpoint()` when `outDir` is set.

## Relation to stratified API

For **single-table stratified** workflows, use `RunCopulaPipeline()` with
`PrepareCopulaData()` and stratum specs. Use `RunFactorVinePipeline()` when you
have separate large prior and small update cohorts.

## Limitations

- Covariate-adjusted marginals: user supplies pre-adjusted data in v1
- Continuous margins only for `FitFactorCopulaPrior()`
- Factor-informed vine truncation uses heuristic `FactorGroupsFromLoadings()`
- Full factor-tree vines deferred

## Session info

```{r session}
sessionInfo()
```
